Hippocampal replay of experience at slow speeds

### 2021-02-27 Article.tex

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{JaiDistincthippocampalcorticalmemory2017}, but another report suggested that events that replay a single location are only seen in young animals \citep{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test. This is problematic because it has the potential to mischaracterize real events that do not move at constant speeds such as those that change direction or are discontinuous \citep{LiuMethodsAssessmentMemory2018}. Furthermore, the use of large, fixed-size temporal bins acts as a boxcar smoother that limits the potential movement speeds of the representation. For example, with 20 ms time bins and 3 cm position bins, an event can only move in 1.5 m/s increments between time steps and all spikes within the 20 ms bins are combined. The linear fit is also highly dependent on the estimation of event boundaries, such as the start and end times of a SWR, because the fit depends on all the data over the course of the event. Approaches that focus on the order of cell activity within each event \citep{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches exclude events that have stationary representations of a single location \citep{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

@@

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (\FIG{1}C). We do this by defining a state transition matrix that defines how the latent position can move from the previous time step (in our case, 2 ms). Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \citep{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, continuous \citep{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \citep{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

@@

Finally, we specify how likely each movement dynamic is to persist in time versus change to another dynamic (\FIG{1}D) via a state transition matrix between dynamics. In order to be conservative with respect to switching between dynamics, we assume that each movement dynamic is likely to dominate for \texttildelow100 ms on average, which is approximately the duration of a SWR event. There is, however, a small probability of switching to one of the other movement dynamics. Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step, which corresponds to an expected duration of 100 ms for staying in a particular dynamic (the Markov assumption of the model means that the probability of staying in a dynamic follows a geometric distribution). Importantly, the data drives the estimated dynamic, so even if the probability of a particular movement dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of \texttildelow10 ms, drive a change to a different estimated dynamic, as in our simulated example. In line with this, note that below we show that our results are relatively insensitive to the value of this parameter.

@@

The second most prevalent classification was exclusively stationary events (4,858 of 21,433 or 23\% classified SWRs). Unlike the stationary-continuous mixtures, most of these events represented a location close to the animal's actual position (\FIG{5}C). There were, however, a number of stationary events that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify non-local stationary events and found such content in 1,437 of 21,433 (\texttildelow7\%) of classified SWRs. 46\% of these (664 of 1,437 stationary SWRs) were stationary throughout the duration of the SWR. These non-local stationary events most commonly represented reward wells or choice points (\FIG{5}E), consistent with \citep{JaiDistincthippocampalcorticalmemory2017}, but a small portion of them occurred at other locations of the track. This suggests that these representations can be flexibly deployed to represent discrete locations.

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (100 cm/s), which is consistent with the running speed of rats (citation?). The next most common category were stationary events \citep{JaiDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated an extended representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events (\texttildelow50-100 ms) with faster continuous dynamics, consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics could also represent either the animal's current location or locations far from the animal, and similarly, previous work has shown that events with continuous dynamics most often begin at the animal's current location \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

@@

We hypothesize that this large set of events supports the ability to rapidly retrieve representations of individual places and snippets of real-world experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{JaiDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to evaluate the value of specific locations or short movement trajectories without having to generate full trajectories to or through them \citep{YuHippocampalcorticalinteraction2015}.

@@

Our identification of the rich and varied dynamics in SWRs depended on several factors. First, we sought to describe as many SWRs as possible, instead of setting a high threshold for event detection. This more permissive approach was also critical for the identification of events that activate specific, immobility-associated locations in previous work \citep{JaiDistincthippocampalcorticalmemory2017}. More broadly, there is no evidence that SWR and high population activity events constitute clearly distinct classes of events that can be identified with a single fixed threshold. Instead, the sizes of these events are well described by a unimodal, long-tailed distribution \citep{ChengNewExperiencesEnhance2008}. High thresholds can therefore exclude large numbers of real events and potentially lead to erroneous conclusions.

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% of values of the posterior probability \citep{CasellaStatisticalinference2001}. In this manuscript, we use the HPD size to evaluate the uncertainty of the posterior probability of position. We calculate the 95\% HPD region by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD region is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The HPD size is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Riemann sum.

\subsection*{Distance from animal}
The distance of the decoded position from the animal's is defined as the shortest path distance along the track graph between the animal's 2D position projected on to the track graph (see Linearization) and the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is defined as the center of the position bin with the greatest posterior value. The shortest path is found using Dijkstra's algorithm \citep{Dijkstranotetwoproblems1959} as implemented in NetworkX \citep{HagbergExploringNetworkStructure2008}.

\subsection*{Estimation of Speed}
In order to estimate the speed over time, we first found the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is defined as the center of the position bin with the greatest posterior value. We then computed the first derivative using the Python library Numpy's gradient function, which computes the difference in the forward and backward time direction and then averages them for all points except the boundaries. We then smoothed this quantity with a small Gaussian (2.5 ms standard deviation) and then averaged this for the classified time points. Because our position bin size of 3 cm makes it hard to distinguish between slower speeds in a 2 ms time step, we only analyzed classifications longer than 20 ms.

>>>>

Interestingly, there is evidence for SWRs where the spiking corresponds to a single location. Specifically, some SWRs contain spiking of neurons associated with single locations where animals are immobile \citep{YuDistincthippocampalcorticalmemory2017}, but another report suggested that events that replay a single location are only seen in young animals \citep{StellaHippocampalReactivationRandom2019}. Thus, while it would seem useful to replay memories associated with single locations and perhaps even experiences occurring at real-world speeds, the prevalence of that type of replay in awake animals remains unclear.

Our uncertainty stems in part from the dominant approaches used to identify the content of replay events. These approaches typically involve multiple steps and assumptions about the nature of replay, which are most commonly characterized using the standard "Bayesian" decoder. First, an encoding model is constructed based on spiking activity during movement, most often using only spikes that have been clustered as single units (putative single neurons). Then, a subset of SWRs or events with high activity levels are selected based on a threshold for event size chosen by the experimenter \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009, StellaHippocampalReactivationRandom2019}. A decoding algorithm is then applied to the spikes within these events, yielding a set of position probability distributions for each time bin. Current approaches use either overlapping or non-overlapping time bins whose size is also chosen by the experimenter. Finally, the most commonly used approach for detecting sequential replay involves fitting a line to the resulting set of probability distributions, which relies on the assumption that the representations progress at a constant, non-zero speed \citep{FosterReversereplaybehavioural2006, DibaForwardreversehippocampal2007a, KarlssonAwakereplayremote2009}. A statistical test is then used to determine whether the line fit to the data is better than a line fit to shuffled versions of the data, where the shuffled version of the data represent some definition of a "random" sequence.

While the standard decoder approach identifies constant speed events, it does not consider events that are rejected by the statistical test. This is problematic because it has the potential to mischaracterize real events that do not move at constant speeds such as those that change direction or are discontinuous \citep{LiuMethodsAssessmentMemory2018}. Furthermore, the use of large, fixed-size temporal bins acts as a boxcar smoother that limits the potential movement speeds of the representation. For example, with 20 ms time bins and 3 cm position bins, an event can only move in 1.5 m/s increments between time steps and all spikes within the 20 ms bins are combined. The linear fit is also highly dependent on the estimation of event boundaries, such as the start and end times of a SWR, because the fit depends on all the data over the course of the event. Approaches that focus on the order of cell activity within each event \citep{LeeMemorySequentialExperience2002, GuptaHippocampalReplayNot2010} allow for a relaxation of that linear assumption, but replace it with a loss of statistical power due to either ignoring all but the first spike from each cell or using an arbitrarily smoothed version of the spike train. These approaches also do not provide information about the dynamics of the underlying spatial representation. Moreover, these approaches often exclude events that have stationary representations of a single location \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}.

@@

Next, we specify movement dynamic models for how latent position---the "mental" position of the animal represented by the cells---evolves over time for a given dynamic (\FIG{1}C). We do this by defining a state transition matrix that defines how the latent position can move from the previous time step (in our case, 2 ms). Previous findings suggest that replay may exhibit at least three distinct types of movement dynamics: stationary \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019}, continuous \citep{DavidsonHippocampalReplayExtended2009}, and fragmented, which could correspond to both extended spatially incoherent representations and representations that jump from one position to another in a short time period \citep{PfeifferAutoassociativedynamicsgeneration2015}. We therefore define state transition models to capture each of these movement dynamics.

@@

Finally, we specify how likely each movement dynamic is to persist in time versus change to another dynamic (\FIG{1}D) via another state transition matrix between the movement dynamics. In order to be conservative with respect to switching between dynamics, we assume that each movement dynamic is likely to dominate for \texttildelow100 ms on average, which is approximately the duration of a SWR event. There is, however, a small probability of switching to one of the other movement dynamics. Accordingly, we set the probability of staying in a dynamic to 0.98 for each 2 ms time step, which corresponds to an expected duration of 100 ms for staying in a particular dynamic (the Markov assumption of the model means that the probability of staying in a dynamic follows a geometric distribution). Importantly, the data drives the estimated dynamic, so even if the probability of a particular movement dynamic is high, data providing clear evidence for a change in dynamic can, within the time frame of \texttildelow10 ms, drive a change to a different estimated dynamic, as in our simulated example. In line with this, note that below we show that our results are relatively insensitive to the value of this parameter.

@@

The second most prevalent classification was exclusively stationary events (4,858 of 21,433 or 23\% classified SWRs). Unlike the stationary-continuous mixtures, most of these events represented a location close to the animal's actual position (\FIG{5}C). There were, however, a number of stationary events that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify non-local stationary events and found such content in 1,437 of 21,433 (\texttildelow7\%) of classified SWRs. 46\% of these (664 of 1,437 stationary SWRs) were stationary throughout the duration of the SWR. These non-local stationary events most commonly represented reward wells or choice points (\FIG{5}E), consistent with \citep{YuDistincthippocampalcorticalmemory2017}, but a small portion of them occurred at other locations of the track. This suggests that these representations can be flexibly deployed to represent discrete locations.

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (100 cm/s), which is consistent with the running speed of rats (citation?). The next most common category were stationary events \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated an extended representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events (\texttildelow50-100 ms) with faster continuous dynamics, consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics could also represent either the animal's current location or locations far from the animal, and similarly, previous work has shown that events with continuous dynamics most often begin at the animal's current location \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

@@

We hypothesize that this large set of events supports the ability to rapidly retrieve representations of individual places and snippets of real-world experience. Previous findings demonstrated that SWRs with movement- or immobility-related activity engage specific patterns of spiking activity outside the hippocampus \citep{YuDistincthippocampalcorticalmemory2017, JadhavCoordinatedExcitationInhibition2016}. Thus, we would expect that the large set of events with slow dynamics would also engage specific patterns of activity outside the hippocampus, perhaps allowing animals to evaluate the value of specific locations or short movement trajectories without having to generate full trajectories to or through them \citep{YuHippocampalcorticalinteraction2015}.

@@

Our identification of the rich and varied dynamics in SWRs depended on several factors. First, we sought to describe as many SWRs as possible, instead of setting a high threshold for event detection. This more permissive approach was also critical for the identification of events that activate specific, immobility-associated locations in previous work \citep{YuDistincthippocampalcorticalmemory2017}. More broadly, there is no evidence that SWR and high population activity events constitute clearly distinct classes of events that can be identified with a single fixed threshold. Instead, the sizes of these events are well described by a unimodal, long-tailed distribution \citep{ChengNewExperiencesEnhance2008}. High thresholds can therefore exclude large numbers of real events and potentially lead to erroneous conclusions.

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% of values of the posterior probability \citep{CasellaStatisticalinference2001}. By using only the top values, this measurement of spread is not influenced by multimodal distributions (whereas an alternative measure like the quantiles of the distribution would be). In this manuscript, we use the HPD size---the total area of the track covered by the 95\% HPD region---to evaluate the uncertainty of the posterior probability of position.

In order to calculate the HPD size, we first calculate the 95\% HPD region by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD region is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The HPD size is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Riemann sum.

\subsection*{Distance from animal}
The distance of the decoded position from the animal's is defined as the shortest path distance along the track graph between the animal's 2D position projected on to the track graph (see Linearization) and the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is the center of the position bin with the greatest posterior value. The shortest path is found using Dijkstra's algorithm \citep{Dijkstranotetwoproblems1959} as implemented in NetworkX \citep{HagbergExploringNetworkStructure2008}.

\subsection*{Estimation of Speed}
In order to estimate the speed over time, we first found the MAP estimate of the posterior probability of position, $\argmax_{x_k} p(x_{k} \mid O_{1:T})$, which is the center of the position bin with the greatest posterior value. We then computed the first derivative using the Python library Numpy's gradient function, which computes the difference in the forward and backward time direction and then averages them for all points except the boundaries. We then smoothed this quantity with a small Gaussian (2.5 ms standard deviation) and then averaged this for the classified time points. Because our position bin size of 3 cm makes it hard to distinguish between slower speeds in a 2 ms time step, we only analyzed classifications longer than 20 ms.
